From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-06 Pi+/paper/reversible1.tex
unifying notation between sec 2 and sec 3; working on small changes to sec 3

\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be specified
by mapping each type to a finite set and each combinator to a permutation between finite sets. In this section, we
outline a simple denotational semantics expressed in the metalanguage of sets and functions that collapses the level-2
combinators. This semantics will be extended in the next two sections to the metalanguage of homotopy type theory to
properly model the level-2 combinators.

@@

\begin{center}
{\scalebox{\scalef}{$%
\begin{array}{rcl}
  \denot{\zerot}     & = & \emptyset                  \\
  \denot{\onet}      & = & \{ \Acon{tt} \}            \\
  \denot{A + B}      & = & \denot{A} \uplus \denot{B} \\
  \denot{A \times B} & = & \denot{A} \times \denot{B}
\end{array}$}}
\end{center}

\noindent where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
sets. The denotation of a combinator $c : A \isoone B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:

\begin{multicols}{2}
%%  \begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
{\scalebox{\scalef}{$%
\begin{array}[t]{rlcl}
  \denot{\identlp}     &      (\inlv{v})           & = & v                                   \\
  \denot{\identrp}     &      v                    & = & \inlv{v}                            \\
  \denot{\swapp}       &     (\inlv{v})           & = & \inrv{v}                            \\
  \denot{\swapp}       &      (\inrv{v})           & = & \inlv{v}                            \\
  \denot{\assoclp}     &      (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
  \denot{\assoclp}     &      (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
  \denot{\assoclp}     &      (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
  \denot{\assocrp}     &      (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
  \denot{\assocrp}     &      (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
  \denot{\assocrp}     &      (\inrv{v})           & = & \inrv{(\inrv{v})}
\end{array}$}}

{\scalebox{\scalef}{$%
\begin{array}[t]{rlcl}
  \denot{\identlt}      &      (\Acon{tt} , v)      & = & v                                   \\
  \denot{\identrt}      &      v                    & = & (\Acon{tt} , v)                     \\
  \denot{\swapt}        &      (v_1 , v_2)          & = & (v_2 , v_1)                         \\
  \denot{\assoclt}     &      (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
  \denot{\assocrt}     &      ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
  \denot{\dist}       &      (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
  \denot{\dist}        &      (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
  \denot{\factor}       &      (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
  \denot{\factor}      &      (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)
\end{array}$}}

@@

\begin{center}
{\scalebox{\scalef}{$%
\begin{array}{rlcl}
  \denot{\idc}         &      v                    & = & v                                   \\
  \denot{(c_1 \fatsemi c_2)} & v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
  \denot{(c_1 \oplus c_2)}   & (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
  \denot{(c_1 \oplus c_2)}   & (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
  \denot{(c_1 \otimes c_2)}  & (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
  \end{array}
$}}
\end{center}

@@

\begin{theorem}\label{thm:semone}
  The denotational semantics is sound in the following sense:
  \begin{itemize}
    \item For every level-1 combinator $c : A \isoone B$, we have that $\denot{c}$ is a bijection between $\denot{A}$ and $\denot{B}$.
    \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
        combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using
        extensional equivalence of functions.
  \end{itemize}
\end{theorem}
\begin{proof}
  For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
  the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and
  $\denot{!c}$ is an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
  $\denot{\AgdaBound{c}~\AgdaOperator{\AgdaInductiveConstructor{◎}}~\AgdaInductiveConstructor{id⟷₁}}
  = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
\end{proof}

>>>>

\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be specified
by mapping each type to a finite set and each combinator to a permutation between finite sets. In this section, we
outline a simple denotational semantics expressed in the metalanguage of sets and functions that collapses the level-2
combinators. This semantics will be extended in the next two sections to the HoTT metalanguage to properly model the
level-2 combinators.

@@

\begin{center}
{\scalebox{\scalef}{$%
\begin{array}{rcl}
  \denot{\zerot}     & = & \bot                  \\
  \denot{\onet}      & = & \top \\
  \denot{A + B}      & = & \denot{A} \sqcup \denot{B} \\
  \denot{A \times B} & = & \denot{A} \times \denot{B}
\end{array}$}}
\end{center}

\noindent where $\bot$ is the empty set, $\top$ is a singleton set, $\sqcup$ is the disjoint union of sets, and $\times$
is the cartesian product of sets. The denotation of a combinator $c : A \isoone B$ is a (bijective) function mapping
$\denot{A}$ to $\denot{B}$:

\begin{multicols}{2}
%%  \begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
{\scalebox{\scalef}{$%
\begin{array}[t]{rlcl}
  \denot{\identlp}     &      (\inl{v})           & = & v                                   \\
  \denot{\identrp}     &      v                    & = & \inl{v}                            \\
  \denot{\swapp}       &     (\inl{v})           & = & \inr{v}                            \\
  \denot{\swapp}       &      (\inr{v})           & = & \inl{v}                            \\
  \denot{\assoclp}     &      (\inl{v})           & = & \inl{(\inl{v})}                   \\
  \denot{\assoclp}     &      (\inr{(\inl{v})})  & = & \inl{(\inr{v})}                   \\
  \denot{\assoclp}     &      (\inr{(\inr{v})})  & = & \inr{v}                            \\
  \denot{\assocrp}     &      (\inl{(\inl{v})})  & = & \inl{v}                            \\
  \denot{\assocrp}     &      (\inl{(\inr{v})})  & = & \inr{(\inl{v})}                   \\
  \denot{\assocrp}     &      (\inr{v})           & = & \inr{(\inr{v})}
\end{array}$}}

{\scalebox{\scalef}{$%
\begin{array}[t]{rlcl}
  \denot{\identlt}      &      (\ttt , v)      & = & v                                   \\
  \denot{\identrt}      &      v                    & = & (\ttt , v)                     \\
  \denot{\swapt}        &      (v_1 , v_2)          & = & (v_2 , v_1)                         \\
  \denot{\assoclt}     &      (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
  \denot{\assocrt}     &      ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
  \denot{\dist}       &      (\inl{v_1} , v_3)   & = & \inl{(v_1 , v_3)}                  \\
  \denot{\dist}        &      (\inr{v_2 , v_3})   & = & \inr{(v_2 , v_3)}                  \\
  \denot{\factor}       &      (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)                  \\
  \denot{\factor}      &      (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)
\end{array}$}}

@@

\begin{center}
{\scalebox{\scalef}{$%
\begin{array}{rlcl}
  \denot{\idc}         &      v                    & = & v                                   \\
  \denot{(c_1 \fatsemi c_2)} & v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
  \denot{(c_1 \oplus c_2)}   & (\inl{v})           & = & \inl{(\denot{c_1}~v)}              \\
  \denot{(c_1 \oplus c_2)}   & (\inr{v})           & = & \inr{(\denot{c_2}~v)}              \\
  \denot{(c_1 \otimes c_2)}  & (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
  \end{array}
$}}
\end{center}

@@

\begin{theorem}\label{thm:semone}
  The denotational semantics is sound in the following sense:
  \begin{itemize}
    \item For every level-1 combinator $c : A \isoone B$, we have that $\denot{c}$ is a bijection between $\denot{A}$ and $\denot{B}$.
    \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
        combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using
        extensional equivalence of functions.
  \end{itemize}
\end{theorem}
\begin{proof}
  For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
  the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and
  $\denot{!c}$ form an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
  $\denot{\AgdaBound{c}~\AgdaOperator{\AgdaInductiveConstructor{◎}}~\AgdaInductiveConstructor{id⟷₁}}
  = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
\end{proof}

### 2021-07-06 Pi+/paper/ufin.tex
messing with packages; missing font problem first reported around p.12 grep code *.tex
unifying notation between sec 2 and sec 3; working on small changes to sec 3

In the previous section, we gave a sound denotational semantics for $\PiLang$ using the groupoid of finite sets and
bijections $\BFin$. Types are interpreted as objects in $\BFin$, and 1-combinators are interpreted as isomorphisms in
$\BFin$. We also showed that if there is a 2-combinator between two 1-combinators, then their interpretations are
extensionally equal in $\BFin$.

However, we want to go a step further, and claim completeness for this semantics, that is, we want to quote back from
the semantics to the syntax. Given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted 1-combinators.

The groupoid $\BFin$ that we presented is strict, that is, two equal permutations are extensionally identified, without
an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$. In a weak category, equational
axioms of the category are replaced by higher invertible cells. For example, there are higher invertible cells for the
left and right units of composition, and associativity. The higher cells come with their own coherence data as well,
such as vertical composition, and horizontal composition, or whiskering.\todo{define 2-category in appendix.}

HoTT provides a rich internal language for working with weak higher groupoids, using the ``Types are weak
$\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has points for 0-cells,
1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible 1-cells.~\footnote{This
is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that
is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type $\UFin$, which is a 1-groupoid. In this
section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT.

@@

  This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
  level~\cite{DBLP:conf/esop/CaretteS16}.
  As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
  we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
  set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
  the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
  $\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
  with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
  \[\begin{array}{rcl}
      \denot{\onet + (\onet + \onet)}                                         & = & \{ 0,1,2 \} \\
      \denot{(\onet + \onet) + \onet}                                         & = & \{ 0,1,2 \} \\
      \\
      \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} & = & (0~1~2)     \\
      \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet}   & = & (2~0~1)
    \end{array}\]
  where we have used the one-line notation for permutations with $(a~b~c)$ representing the
  permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
  enumeration of the elements of each type:
  \[\begin{array}{rcl}
      \mathit{enum}(\zerot)     & = & [~]                                                                                                          \\
      \mathit{enum}(\onet)      & = & [ ~\Acon{tt}~ ]                                                                                              \\
      \mathit{enum}(A + B)      & = & \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B)          \\
      \mathit{enum}(A \times B) & = & \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
    \end{array}\]
  \noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
  a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
  operation that appends all the subsequences in a sequence.

>>>>

The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are sets and
whose (iso)morphisms as bijections. The semantics interprets types as objects in $\BFin$, 1-combinators as isomorphisms
in $\BFin$, and for every pair of 1-combinators related by a 2-combinator, their interpretations are extensionally equal
in $\BFin$. This groupoid is strict: two equal permutations are extensionally identified, without an explicit witness
for the equality. In order to establish completeness, we want to quote back from the semantics to the
syntax. Specifically, given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted
1-combinators. Our first step is therefore to \emph{weaken} this groupoid $\BFin$.

In a weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are
higher invertible cells for the left and right units of composition, and associativity. The higher cells come with their
own coherence data as well, such as vertical composition, and horizontal composition, or whiskering.\todo{define
  2-category in appendix.} HoTT provides a rich internal language for working with weak higher groupoids, using the
``Types are weak $\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has
points for 0-cells, 1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible
1-cells.~\footnote{This is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the
  hom-categories are posets, that is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type
$\UFin$, which is a 1-groupoid. In this section, we develop the tools necessary to construct $\UFin$, starting with a
quick review of HoTT.

@@

  This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
  level~\cite{carette2016}.
  As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
  we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
  set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
  the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
  $\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
  with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
  \[\begin{array}{rcl}
      \denot{\onet + (\onet + \onet)}                                         & = & \{ 0,1,2 \} \\
      \denot{(\onet + \onet) + \onet}                                         & = & \{ 0,1,2 \} \\
      \\
      \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} & = & (0~1~2)     \\
      \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet}   & = & (2~0~1)
    \end{array}\]
  where we have used the one-line notation for permutations with $(a~b~c)$ representing the
  permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
  enumeration of the elements of each type:
  \[\begin{array}{rcl}
      \mathit{enum}(\zerot)     & = & [~]                                                                                                          \\
      \mathit{enum}(\onet)      & = & [ ~\Acon{tt}~ ]                                                                                              \\
      \mathit{enum}(A + B)      & = & \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B)          \\
      \mathit{enum}(A \times B) & = & \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
    \end{array}\]
  \noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
  a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
  operation that appends all the subsequences in a sequence.
